Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes

### 2024-03-05 manuscript/main-revision.tex
Added figures for revision

% Research Article (7500 words)
\section*{Introduction}\label{sec:introduction}

@@

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

@@

\subsubsection*{Within-species trait variations}

@@

\subsubsection*{Between-species trait variations}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

\subsubsection*{Neutrality index}

@@

\subsection*{Estimation}\label{subsec:estimate}

@@

\subsubsection*{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitutions per site) and the topology of the phylogenetic tree.
The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

@@

\subsubsection*{Multivariate Brownian process}

@@

\subsubsection*{Bayesian estimate}
The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\parencite{latrille_inferring_2021}.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\EstRateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\EstNI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

\subsubsection*{Applicability to empirical data}

@@

\subsection*{Simulation}\label{subsec:simulations}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
    \caption{
        Wright-Fisher simulations with mutation, selection and drift.
        Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
        The trait's genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRatePheno$ to mutate.
        Drift is modeled by the resampling of parents.
        Right panel: examples of a trait evolving along a phylogenetic tree, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
    }
    \label{fig:simulator}
\end{figure*}

\subsection*{Empirical dataset}\label{subsec:empirical-dataset}

@@

\section*{Results}\label{sec:results}

\subsection*{Neutrality index}

@@

\subsection*{Results against simulations}\label{subsec:results-against-simulations}

@@

\begin{figure*}[!ht]
    \centering
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.rho}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.pvalues}
    \end{minipage}
    \caption{
        $10,000$ simulations of trait evolution along a phylogenetic tree under different selection regimes.
        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
        Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
        Effective population size ($\Ne$) and mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
    }
    \label{fig:results-simulations}
\end{figure*}

@@

\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}

@@

\section*{Discussion}\label{sec:discussion}

@@

\begin{center}
    \captionof{figure}{summary of trait's genetic architecture.}
    \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
    \label{fig:simulator-summary}
\end{center}

>>>>

% Research Article (7500 words)
\section{Introduction}\label{sec:introduction}

@@

\section{Materials and Methods}\label{sec:materials-and-methods}
\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

@@

\subsubsection{Within-species trait variations}

@@

\subsubsection{Between-species trait variations}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure1}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

\subsubsection{Neutrality index}

@@

\subsection{Estimation}\label{subsec:estimate}

@@

\subsubsection{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitutions per site) and the topology of the phylogenetic tree.
The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

@@

\subsubsection{Multivariate Brownian process}

@@

\subsubsection{Bayesian estimate}
The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\EstNI$) for traits of interest.
We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\parencite{latrille_inferring_2021}.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\EstRateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\EstNI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

\subsubsection{Applicability to empirical data}

@@

\subsection{Simulation}\label{subsec:simulations}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure2}
    \caption{
        Wright-Fisher simulations with mutation, selection and drift.
        Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
        The trait's genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRatePheno$ to mutate.
        Drift is modeled by the resampling of parents.
        Right panel: examples of a trait evolving along a phylogenetic tree, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
    }
    \label{fig:simulator}
\end{figure*}

\subsection{Empirical dataset}\label{subsec:empirical-dataset}

@@

\section{Results}\label{sec:results}

\subsection{Neutrality index}

@@

\subsection{Results against simulations}\label{subsec:results-against-simulations}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure3}
    \caption{
        $10,000$ simulations of trait evolution along a phylogenetic tree under different selection regimes.
        Traits simulated under stabilizing selection (blue), under a neutral evolution (yellow), and under a moving optimum (red).
        Histogram of ratio of between-species trait variation ($\EstRateBetween$) over within-species trait variation $\EstRateWhithin$ with $\EstNI = \EstRateBetween / \EstRateWhithin$ estimated from each simulated data (left) and probabilities of $\EstNI$ being greater than $1$ (right).
        Effective population size ($\Ne$) and mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$) were either constant (top row), or fluctuating as a Brownian process along the phylogenetic tree (bottom row).
    }
    \label{fig:results-simulations}
\end{figure*}

@@

\subsection{Results on empirical data}\label{subsec:results-on-empirical-data}

@@

\section{Discussion}\label{sec:discussion}

@@

\begin{center}
    \captionof{figure}{summary of trait's genetic architecture.}
    \includegraphics[width=0.6\textwidth, page=1] {figureS1}
    \label{fig:simulator-summary}
\end{center}
